---
title             : "The effect of erroneous R code on student performance"
shorttitle        : "Erroneous R code"

author: 
  - name          : "James Bartlett"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "62 Hillhead Street, Glasgow"
    email         : "james.bartlett@glasgow.ac.uk"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - "Conceptualization"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"

affiliation:
  - id            : "1"
    institution   : "University of Glasgow, United Kingdom"

authornote: |
  Enter author note here.

abstract: |
  Blah blah blah. 
  
keywords          : "learning, R, statistical programming, error-full learning"
wordcount         : "1200"

bibliography      : ["references.bib", "r-references.bib"]
csl               : apa7.csl

floatsintext      : no # 
linenumbers       : yes # Add line numbers? 
draft             : no # Add draft watermark on every page? 
mask              : no # Hide author details for blind submission? 

figurelist        : no # Add figure list? 
tablelist         : no # Add table list? 
footnotelist      : no # Add footnote list? 

classoption       : "man"
output            : papaja::apa6_pdf # Can change ending to _word to knit to Word
---

```{r setup, include = FALSE, message=FALSE, warning=FALSE}
library(papaja)
library(pwr)
library(tidyverse)

r_refs("r-references.bib")
```

```{r analysis-preferences}
# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
```

Data skills are increasingly recognised as a key component of psychological literacy. To promote reproducible data preparation and analysis workflows, educators have highlighted the role of teaching students how to use statistical programming languages instead of point-and-click software [@mcaleer_embedding_2022]. However, programming is rare in UK psychology curricula [@targ_meta-research_group_statistics_2022] and offers unique challenges such as how to prepare students to debug their code. Debugging code is a separate problem solving skill to learn alongside statistics, so it is important to understand how best to teach students debugging skills. 

@hoffman_students_2021 reported a small pilot study using SAS where they compared a traditional error-free course structure to an error-full course focusing on debugging errors alongside key concepts. 80% of students preferred the error-full course but the study only included 18 participants and just 4 students completed assignments following each course, meaning they could not compare performance. Therefore, in our study, we want to apply these methods to the programming language R and recruit a larger sample. 

We hypothesise that students who complete the error-full lecture will score higher on a data skills assignment than students who complete the error-free lecture. 

# Methods

## Participants

## Sample size justification

```{r}
pwr.t2n.test(n1 = 25, n2 = 50, power = 0.8, sig.level = 0.5)


```


## Material

## Procedure

## Data analysis
We used `r cite_r("r-references.bib")` for all our analyses.


# Results

# Discussion


\newpage

# References

::: {#refs custom-style="Bibliography"}
:::
